Titeln på avhandlingens språk

### 2022-08-31 publications/2021ImpactDelayedResponse/Manuscript.tex
plos

% BibLatex and PLOS One hacks
\usepackage[citestyle=numeric,
            backend=bibtex,
            bibstyle=authoryear,
            sorting=none,
            sortcites,
            firstinits,
            terseinits,
            maxbibnames=6,
            dashed=false]{biblatex}
\bibliography{bibliography}
\usepackage[colorlinks]{hyperref}

% Name list format
\renewcommand*{\labelnamepunct}{\addspace}
\renewcommand*{\finalnamedelim}{%
  \ifbibliography{\addcomma\space}{\addspace\&\space}}
\renewcommand*{\revsdnamepunct}{}
\DeclareNameAlias{sortname}{last-first}

% No quotes or italics in titles
\DeclareFieldFormat{title}{#1}
\DeclareFieldFormat[article,inbook,incollection,inproceedings,patent,
                    thesis,unpublished]{title}{#1}
\DeclareFieldFormat{journaltitle}{#1}
\DeclareFieldFormat{issuetitle}{#1}
\DeclareFieldFormat{maintitle}{#1}

\renewbibmacro*{in:}{%
  \ifentrytype{article}{}{\printtext{\bibstring{in}\intitlepunct}}}

% Remove page prefixes
\DeclareFieldFormat{pages}{#1}

% Only print journal volume
\renewbibmacro*{volume+number+eid}{%
  \setunit*{\addspace}%
  \printfield{volume}}

% Use colon for volume-pages delimiter
\renewcommand*{\bibpagespunct}{%
  \ifentrytype{article}{\addcolon\space}{\addcomma\space}}

% Add labelnumbers to bibliography
\DeclareFieldFormat{labelnumberwidth}{#1\adddot}
\defbibenvironment{bibliography}
  {\list
     {\printtext[labelnumberwidth]{%
      \printfield{prefixnumber}%
      \printfield{labelnumber}}}
     {\setlength{\labelwidth}{\labelnumberwidth}%
      \setlength{\leftmargin}{\labelwidth}%
      \setlength{\labelsep}{\biblabelsep}%
      \addtolength{\leftmargin}{\labelsep}%
      \setlength{\itemsep}{\bibitemsep}%
      \setlength{\parsep}{\bibparsep}}%
      \renewcommand*{\makelabel}[1]{\hss##1}}
  {\endlist}
  {\item}

% Omit authoryear disambiguation
\AtEveryBibitem{\clearfield{extrayear}}

@@

Wearable Cognitive Assistants, WCA for short, have recently started to garner attention from the research community~\autocite{Ha:TowardsWearableCogAssist,Chen:EarlyImplementation}.
They represent a novel category of highly interactive and context-sensitive augmented reality applications, that aim to amplify human cognition in both day-to-day tasks and professional settings.
Their mode of operation is analogous to how GPS navigation systems guide drivers, by seamlessly providing relevant instructions and feedback relating to the current task at hand.
Note that this implies seamless interaction with the context of the user --- at no moment does the user need to trigger an update explicitly, as the application is constantly tracking the state of the target task.
An example is an \textcite{IKEAAssistant} which monitors the assembly of a piece of furniture in real time, providing timely, step-by-step feedback to guide the user toward completion.

WCA systems were originally inspired by assistive use cases for people suffering from some form of cognitive decline, either through aging or because of traumatic brain injuries~\autocite{Ha:TowardsWearableCogAssist,satya2019augmenting}.
More recently, they have been applied to a broader range of use cases, including step-by-step guidance on complex assembly tasks~\autocite{Chen:AnEmpiricalStudyOfLatency}.
Non-wearable augmented reality and cognitive assistance systems have already been proven to be valuable tools in the industrial workplace~\autocite{funk2015caworkplace,gorecky2011cognito}.
Detethering this assistance from its current fixed location will surely make it available to many more fields.

@@

    \item\label{item:lowlatency} Feedback should be ``quick'', relative to the task at hand. 
    This requirement is further strengthened by the previously mentioned ``seamless interaction'' characteristic of these systems.
    This means that users will have expectations of constant, immediate feedback as they progress through the task.
    In the case of a step-by-step task like IKEA, delayed feedback might simply confuse or distract the user.
    However, in a highly interactive task like a \emph{Ping-Pong assistant}~\autocite{PingPongAssistant, Chen:EarlyImplementation} late guidance is at best a nuisance and at worst a severe handicap.
\end{enumerate}

@@

% However, additional challenges remain in the path toward widespread adoption.
\cref{item:lowlatency} forms the base motivation for the research presented in this paper.
We still have a very limited understanding of how humans react to delays in these systems --- specifically, how changes in system responsiveness impact users.
\emph{System responsiveness} here denotes a qualitative scale ranging from \emph{high} (that is, not subject to delay or subject to negligible delay with respect to human perception) to \emph{low} (i.e.\ subject to considerable delay).
% Conversely, QoE is understood to range from \emph{good} (i.e.\ users enjoy interacting with the application) to bad (users dislike interacting with the application).
Characterizing the relationships between system responsiveness and user behavior and quality of experience is of paramount importance for the design and evaluation of these applications.
It is generally acknowledged, for instance, that a system going from a state of high responsiveness to one of low responsiveness can cause a drop in quality of experience~\autocite{dabrowsky:2011:40years}. 
Furthermore, this could cause users to modify the temporal paramaters of their behavior when interacting with the system, generating a sort of \emph{feedback loop} between user and system.
A clear understanding of these relationships would allow, for instance, for the development of strategies for load balancing and optimization for large-scale deployment of WCA.\@
% Although maximizing QoE through over-engineered systems and networks would be a ``simple'' solution to this problem, in practice, the prohibitive costs of infrastructure make finding the optimal trade-off between QoE and cost a valuable endeavor.

This paper builds upon preliminary work in the field of time perception and delay characterization of WCA.\@
We expand upon the findings of \textcite{Ha:TowardsWearableCogAssist}, who identified the need for low-latency offloading in WCA, and of \textcite{Chen:AnEmpiricalStudyOfLatency}, who outlined the bounds for ``noticeable'' and ``unbearable'' latencies in these systems.
While these bounds present a general understanding of when it is likely that a user will stop using an application, they do not provide any insights as to what happens \emph{before} that --- i.e.\ how human behavior changes with system responsiveness.
We aim to tackle this gap in knowledge through the characterization of human responses to delays in the application pipeline, using latencies in the range defined by the previously established bounds.
This is an important step toward a more systematic understanding of human behavior in this domain.

We present in this paper the design and elaboration of an experimental WCA test-bed.
This test-bed was subsequently employed in a study in which undergraduate students of diverse fields of study, aged between 18 and 25 years old, were asked to interact with and follow the instructions given to them by a cognitive assistant.
Unbeknownst to the participants, we altered the responsiveness of the system in real-time and recorded the resulting behavioral and physiological reactions. 
The participants wore an array of biometric sensors measuring physiological responses that have proven useful in assessing cognitive workload during human-computer interaction~\autocite{haapalainen2010psycho,kumar2016measurement} such as heart rate and EEG.\@

@@

  Finally, we hypothesized that the individual trait of \emph{neuroticism}~\autocite{john1999:bfi} would play a role in mediating these effects, as it has previously been connected to intolerance for time delay~\autocite{hirsh2008delay}.
  We also expected \emph{focus} and \emph{involvement}~\autocite{witmer1998:itq} to play roles in this.

@@

% Our results indicate that reduced responsiveness in WCA systems leads to a disruption of participants' cognitive plan for the task.
% This is evidenced by an emergent pacing effect on user actions as system responsiveness is reduced.
% While it would seem self-evident that users take longer to complete a task while using a system with low responsiveness --- as they have to wait longer for new instructions --- our study found that user slow-down represents a source of substantial additional delay.
% To be more precise, the data indicate that users slow down not only because they have to wait for the system to catch up, but that their reactions to new instructions is also delayed.
% Furthermore, this effect persists for a while after system response improves and is modulated by intrinsic personality traits, in particular, \emph{neuroticism}~\autocite{john1999:bfi}.

@@

Our hope is that the results we provide might prove useful for the understanding and optimization of deployments of WCA.\@
These results represent unexpected, valuable findings, which can be employed to model and understand how users interact with latency in applications and systems, and develop resource allocation and power optimization strategies.
Additionally, we hope that the results we provide might pave the way for the improvement of performance evaluation tools such as our previous work in \textcite{olguin:2018, olguin:2019}.
Such systems would greatly benefit from this knowledge, as it would allow for the design and implementation of realistic models of human behavior, making highly accurate benchmarks a reality in the domain of WCA.\@

@@

% Our overarching theoretical construct is the idea that people perform temporally sequential tasks in the framework of cognition and emotion.
% Additionally, we hypothesize that individual diferences between subjects moderate responses to delays in the feedback during the execution of a task.
% In particular, we view cognition as described by models such as {ACT-R}~\autocite{neves1981knowledge}; i.e.\ as a sequence of procedures performed by the subject.

Our overarching theoretical construct is the idea that people perform temporally sequential tasks in the framework of cognition and emotion.
In particular, we view cognition as described by models such as {ACT-R}~\autocite{neves1981knowledge}; i.e.\ as a sequence of procedures performed by the subject.
Additionally, we hypothesize that individual diferences between subjects moderate responses to delays in the feedback during the execution of a task.

@@

The question of how people respond to delay in a computer system is grounded in how people perceive time.
Time perception has been described as regulated by an attentional gate that, when opened, starts a cognitive pulse counter~\autocite{zakay1995attentional,zakay1996role}.
More recent research indicates, however, that duration perception is highly malleable and the result of multiple timing mechanisms found in overlapping, flexible neural systems~\autocite{bruno2016multiple,wiener2011multiple}. 
The estimation of an event's duration varies with context of various types
\begin{enumerate*}[label={(\roman*)}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
    \item events subsequent to a long or short interval are contracted or extended, respectively~\autocite{heron2012duration}
    \item repeated events tend to be perceived as shorter than novel ones~\autocite{matthews2011stimulus}
    \item arousal can expand durations~\autocite{droit2011emotion}.
\end{enumerate*}

Expectations play a critical role in time perception as well~\autocite{zakay1995attentional,zakay1996role}.
It has been shown that people have a general tendency to be hypersensitive to delays in worse-than-expected states, and under-sensitive to meeting or exceeding expectations~\autocite{Loewenstein1992anomaliesintertemporalchoice}.
Accordingly, failures to meet expected fast response times tend to be experienced as highly negative, whereas fast latencies are not noticed.
Violations of expectancy have a strong impact on the acceptability of computer systems.
Users of a computer system anticipate the latency for events, for which the standards only become more stringent as systems improve in response time.
In immersive systems like WCA, which aim to provide seamless interaction, delays are particularly noticeable.

It has long been recognized that slow system response times can undermine cognitive processing, slow the pace of users, and lead to stereotyped behavior and errors, as well as cause negative emotional consequences~\autocite{dabrowsky:2011:40years}.
However, standards for what constitute tolerable delays have changed dramatically compared to three decades ago, when delays on the order of \SI{10}{\second} were deemed acceptable~\autocite{nielsen1994usability, shneiderman2016designing, seow2008designing}.
Today's user context, and WCA in particular, often demand response times orders of magnitude shorter.

For WCA the acceptable range for latencies was explored by \textcite{Chen:AnEmpiricalStudyOfLatency}, by constructing assistants for tasks with a range of time constraints, including step-by-step tasks and more interactive contexts like playing Ping-Pong against a human opponent.
They then proposed a latency tolerance zone according to the task demands.
For an essentially self-paced task like LEGO assembly, they found two key ranges of latency; unnoticeable, \SIrange{0}{0.6}{\second}; and impaired, \SIrange{0.6}{2.7}{\second}. 
Beyond that, users could begin to show the negative outcomes previously catalogued~\autocite{dabrowsky:2011:40years}.

@@

A first possible explanation comes from research on cognitive and motor planning. 
Delay may move users from relatively automatic to more attention-demanding processing.
Cognitive and motor tasks are commonly described as a hierarchical system, progressing from high-level goals to the sequence of commands that accomplishes them.
As competency in a task increases, execution of the hierarchy becomes increasingly automated.
Automatization has been described from a computational perspective in {Anderson's ACT-R}~model as the compiling of multiple productions into one~\autocite{neves1981knowledge}.
Neural measurements indicate that with automaticity, control moves from frontal brain areas to more posterior ones~\autocite{jeon2015degree,puttemans2005changes}, and similar distinctions have been related to temporal processing~\autocite{lewis2003distinct,koch2009neural,lee2019limiting}.

Although activities guided by a WCA are not simple motor actions, immediate feedback after each of a series of repeated actions should promote development and automatic execution of a hierarchical plan.
Delays, in contrast, would disrupt such a plan through the loss of automated control~\autocite{lee2019limiting}.

A second, alternative explanation of delay effects appeals to emotional systems rather than cognitive processes.
As users of a system become emotionally aroused by delay, they may be subject to generalized arousal, causing decrements in performance~\autocite{lee2019limiting}.

Finally, a third potential explanation of delay effects is what has been called ``ego depletion'', the notion that expending effort on self-control eliminates resources needed for further effort~\autocite{baumeister74tice,lin2020strong}.

The various processing accounts of delay effects predict different outcomes, which we will consider in the context of the current data.
If delay increases attentional demands on cognitive processes, responses should be slowed and errors expected, particularly on time-critical tasks.
Generalized arousal triggered by emotional stress from delay should emerge in physiological measures, such as increased heart rate or skin conductivity.
Arousal can also reduce movement smoothness or add erratic gestures~\autocite{pijpers2003anxiety}.
Ego depletion has been found to produce premature responses culminating in error~\autocite{lin2020strong}, or to lead to abandoning a task entirely~\autocite{baumeister74tice}.

Over-arching prescriptions for tolerable system response time have not tended to take into account individual differences in users with respect to salient variables like cognitive ability or personality. 
Relevant research can be found in studies of delay discounting, the tendency to devalue rewards for which one must wait.
High discounting rates, indicative of waiting intolerance, have been associated with negative social and academic outcomes.
\textcite{hirsh2008delay} found that higher discounting was associated with extraversion among those with low cognitive function, whereas lower discounting was associated with emotional stability (low neuroticism) for people with high cognitive function.
Among computer system users who tend to have relatively high cognitive ability (which presumably describes the present experimental population), this points to neuroticism as a personality factor that might modulate tolerance for waiting. Extraversion could also  be  a moderating factor among the broader target audience of WCA, which are intended for relatively inexperienced users of an application.
These and other measures of individual variation were considered here.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.8\textwidth]{images/experiment_arch.pdf}
  \caption{Experimental test-bed.
  Participants interact with the cognitive assistant through task-related inputs and outputs --- in practice, these correspond to the video feed captured by the assistant and the instructions provided by it.
  The assistant itself has been instrumented with a data collection layer, which collects and processes experiment-related data such as biometric signals from the participants (these are merely processed here and do not form part of the inputs to the cognitive assistant as such, however), and a delay buffer, which introduces controlled delays in the transit of information from the core processing component.}%
  \label{fig:experimentaltestbed}
\end{figure}

This study was conducted with the approval of the Carnegie Mellon University Institutional Review Board.

@@

We used a modified version of the LEGO Assistant application introduced by \textcite{Chen:EarlyImplementation}.
This application belongs to a category of WCAs designed to guide users through the execution of a sequential task.
% Such applications constitute ``\emph{conversational computing tasks}'' in the taxonomy proposed by \textcite{dabrowsky:2011:40years}.
The assistant presents a series of instructions to the user in a semi-predetermined order while it monitors the progress in real-time.
Whenever the application detects that the user has correctly performed an instruction, it provides a new one.
In the case of an incorrect action by the user, a procedurally generated corrective instruction is provided instead.
% The results of the user performing these instructions are provided as inputs to the system.
% These inputs may either be correct, in which case the system proceeds to output the next instruction, or incorrect, in which case a procedurally generated corrective step that fixes the mistake is provided to the user.

@@

\begin{figure*}[h]
  \centering
  \includegraphics[width=.8\textwidth]{images/step_task_description.pdf}
  \caption{Example of a Cognitive Assistance task and its component steps.}\label{fig:task:steps}
\end{figure*}

@@

\begin{figure*}[h]
  \centering
  % \input{plots/lego_hierarchy_diagram.tex}
  \includegraphics[width=\textwidth]{plots/lego_hierarchy_diagram_standalone.eps}
  \caption{Hierarchical cognitive structure of a step in the LEGO task.}
  \label{fig:lego:hierarchical}
\end{figure*}

@@

For the experimental LEGO task, a key modification was made to the structure of the steps.
After each the processing of each input is completed, the result is withheld for a variable period of time until a specific target \emph{delay} is reached, as illustrated in \cref{fig:cogassist:step,fig:cogassist:step:delay}.
The length of this delay is one of two independent variables we manipulated for the experiment.
Seven levels of delay were used --- no added delay (which we will also refer to as \SI{0}{\second} delay), \SIlist{0.6;1.125;1.65;2.175;2.7;3.0}{\second} --- chosen based on the latency bounds found previously by \textcite{Chen:AnEmpiricalStudyOfLatency}.
% A value of  \SI{600}{\milli\second} was identified as the bound where users start noticing delays in the assistant.
% Conversely, \SI{2.7}{\second} was identified as the upper bound on delays after which the application is considered to be in such a degraded state it is basically ``unusable''.
Our selection of delays is centered around the range of delays where latency is noticeable to users, but the application remains in a ``usable'' state.
% while including one delay value in the unnoticeable range and one completely in the ``unusable'' range.

@@

The first of these, the Big Five Inventory of Personality (\emph{BFI}~\autocite{john1999:bfi}), consists of 44 questions to be answered on a 5-point \emph{Likert}-type scale, assessing the traits of \emph{agreeableness: detached to compassionate} (8 questions), \emph{conscientiousness: careless to organized} (9 questions), \emph{extroversion: reserved to outgoing} (8 questions), \emph{neuroticism: secure to sensitive} (8 questions), and \emph{openness: cautious to curious} (9 questions).
Of these, extroversion and neuroticism have been related to tolerance for delay~\autocite{hirsh2008delay}.

>>>>

% BibLatex and PLOS One hacks
% \usepackage[citestyle=numeric,
%             backend=bibtex,
%             bibstyle=authoryear,
%             sorting=none,
%             sortcites,
%             firstinits,
%             terseinits,
%             maxbibnames=6,
%             dashed=false]{biblatex}
% \bibliography{bibliography}
\usepackage{cite}

\usepackage[colorlinks]{hyperref}

% Name list format
% \renewcommand*{\labelnamepunct}{\addspace}
% \renewcommand*{\finalnamedelim}{%
%   \ifbibliography{\addcomma\space}{\addspace\&\space}}
% \renewcommand*{\revsdnamepunct}{}
% \DeclareNameAlias{sortname}{last-first}

% No quotes or italics in titles
% \DeclareFieldFormat{title}{#1}
% \DeclareFieldFormat[article,inbook,incollection,inproceedings,patent,
%                     thesis,unpublished]{title}{#1}
% \DeclareFieldFormat{journaltitle}{#1}
% \DeclareFieldFormat{issuetitle}{#1}
% \DeclareFieldFormat{maintitle}{#1}

% \renewbibmacro*{in:}{%
%   \ifentrytype{article}{}{\printtext{\bibstring{in}\intitlepunct}}}

% Remove page prefixes
% \DeclareFieldFormat{pages}{#1}

% Only print journal volume
% \renewbibmacro*{volume+number+eid}{%
%   \setunit*{\addspace}%
%   \printfield{volume}}

% Use colon for volume-pages delimiter
% \renewcommand*{\bibpagespunct}{%
%   \ifentrytype{article}{\addcolon\space}{\addcomma\space}}

% Add labelnumbers to bibliography
% \DeclareFieldFormat{labelnumberwidth}{#1\adddot}
% \defbibenvironment{bibliography}
%   {\list
%      {\printtext[labelnumberwidth]{%
%       \printfield{prefixnumber}%
%       \printfield{labelnumber}}}
%      {\setlength{\labelwidth}{\labelnumberwidth}%
%       \setlength{\leftmargin}{\labelwidth}%
%       \setlength{\labelsep}{\biblabelsep}%
%       \addtolength{\leftmargin}{\labelsep}%
%       \setlength{\itemsep}{\bibitemsep}%
%       \setlength{\parsep}{\bibparsep}}%
%       \renewcommand*{\makelabel}[1]{\hss##1}}
%   {\endlist}
%   {\item}

% Omit authoryear disambiguation
% \AtEveryBibitem{\clearfield{extrayear}}

@@

Wearable Cognitive Assistants, WCA for short, have recently started to garner attention from the research community~\cite{Ha:TowardsWearableCogAssist,Chen:EarlyImplementation}.
They represent a novel category of highly interactive and context-sensitive augmented reality applications, that aim to amplify human cognition in both day-to-day tasks and professional settings.
Their mode of operation is analogous to how GPS navigation systems guide drivers, by seamlessly providing relevant instructions and feedback relating to the current task at hand.
Note that this implies seamless interaction with the context of the user --- at no moment does the user need to trigger an update explicitly, as the application is constantly tracking the state of the target task.
An example is an IKEA Assistant~\cite{IKEAAssistant} which monitors the assembly of a piece of furniture in real time, providing timely, step-by-step feedback to guide the user toward completion.

WCA systems were originally inspired by assistive use cases for people suffering from some form of cognitive decline, either through aging or because of traumatic brain injuries~\cite{Ha:TowardsWearableCogAssist,satya2019augmenting}.
More recently, they have been applied to a broader range of use cases, including step-by-step guidance on complex assembly tasks~\cite{Chen:AnEmpiricalStudyOfLatency}.
Non-wearable augmented reality and cognitive assistance systems have already been proven to be valuable tools in the industrial workplace~\cite{funk2015caworkplace,gorecky2011cognito}.
Detethering this assistance from its current fixed location will surely make it available to many more fields.

@@

    \item\label{item:lowlatency} Feedback should be ``quick'', relative to the task at hand. 
    This requirement is further strengthened by the previously mentioned ``seamless interaction'' characteristic of these systems.
    This means that users will have expectations of constant, immediate feedback as they progress through the task.
    In the case of a step-by-step task like IKEA, delayed feedback might simply confuse or distract the user.
    However, in a highly interactive task like a \emph{Ping-Pong assistant}~\cite{PingPongAssistant, Chen:EarlyImplementation} late guidance is at best a nuisance and at worst a severe handicap.
\end{enumerate}

@@

% However, additional challenges remain in the path toward widespread adoption.
\cref{item:lowlatency} forms the base motivation for the research presented in this paper.
We still have a very limited understanding of how humans react to delays in these systems --- specifically, how changes in system responsiveness impact users.
\emph{System responsiveness} here denotes a qualitative scale ranging from \emph{high} (that is, not subject to delay or subject to negligible delay with respect to human perception) to \emph{low} (i.e.\ subject to considerable delay).
% Conversely, QoE is understood to range from \emph{good} (i.e.\ users enjoy interacting with the application) to bad (users dislike interacting with the application).
Characterizing the relationships between system responsiveness and user behavior and quality of experience is of paramount importance for the design and evaluation of these applications.
It is generally acknowledged, for instance, that a system going from a state of high responsiveness to one of low responsiveness can cause a drop in quality of experience~\cite{dabrowsky:2011:40years}. 
Furthermore, this could cause users to modify the temporal paramaters of their behavior when interacting with the system, generating a sort of \emph{feedback loop} between user and system.
A clear understanding of these relationships would allow, for instance, for the development of strategies for load balancing and optimization for large-scale deployment of WCA.\@
% Although maximizing QoE through over-engineered systems and networks would be a ``simple'' solution to this problem, in practice, the prohibitive costs of infrastructure make finding the optimal trade-off between QoE and cost a valuable endeavor.

This paper builds upon preliminary work in the field of time perception and delay characterization of WCA.\@
We expand upon the findings of Ha et al.~\cite{Ha:TowardsWearableCogAssist}, who identified the need for low-latency offloading in WCA, and of Chen et al.~\cite{Chen:AnEmpiricalStudyOfLatency}, who outlined the bounds for ``noticeable'' and ``unbearable'' latencies in these systems.
While these bounds present a general understanding of when it is likely that a user will stop using an application, they do not provide any insights as to what happens \emph{before} that --- i.e.\ how human behavior changes with system responsiveness.
We aim to tackle this gap in knowledge through the characterization of human responses to delays in the application pipeline, using latencies in the range defined by the previously established bounds.
This is an important step toward a more systematic understanding of human behavior in this domain.

We present in this paper the design and elaboration of an experimental WCA test-bed.
This test-bed was subsequently employed in a study in which undergraduate students of diverse fields of study, aged between 18 and 25 years old, were asked to interact with and follow the instructions given to them by a cognitive assistant.
Unbeknownst to the participants, we altered the responsiveness of the system in real-time and recorded the resulting behavioral and physiological reactions. 
The participants wore an array of biometric sensors measuring physiological responses that have proven useful in assessing cognitive workload during human-computer interaction~\cite{haapalainen2010psycho,kumar2016measurement} such as heart rate and EEG.\@

@@

  Finally, we hypothesized that the individual trait of \emph{neuroticism}~\cite{john1999:bfi} would play a role in mediating these effects, as it has previously been connected to intolerance for time delay~\cite{hirsh2008delay}.
  We also expected \emph{focus} and \emph{involvement}~\cite{witmer1998:itq} to play roles in this.

@@

% Our results indicate that reduced responsiveness in WCA systems leads to a disruption of participants' cognitive plan for the task.
% This is evidenced by an emergent pacing effect on user actions as system responsiveness is reduced.
% While it would seem self-evident that users take longer to complete a task while using a system with low responsiveness --- as they have to wait longer for new instructions --- our study found that user slow-down represents a source of substantial additional delay.
% To be more precise, the data indicate that users slow down not only because they have to wait for the system to catch up, but that their reactions to new instructions is also delayed.
% Furthermore, this effect persists for a while after system response improves and is modulated by intrinsic personality traits, in particular, \emph{neuroticism}~\cite{john1999:bfi}.

@@

Our hope is that the results we provide might prove useful for the understanding and optimization of deployments of WCA.\@
These results represent unexpected, valuable findings, which can be employed to model and understand how users interact with latency in applications and systems, and develop resource allocation and power optimization strategies.
Additionally, we hope that the results we provide might pave the way for the improvement of performance evaluation tools such as our previous work in \cite{olguin:2018, olguin:2019}.
Such systems would greatly benefit from this knowledge, as it would allow for the design and implementation of realistic models of human behavior, making highly accurate benchmarks a reality in the domain of WCA.\@

@@

% Our overarching theoretical construct is the idea that people perform temporally sequential tasks in the framework of cognition and emotion.
% Additionally, we hypothesize that individual diferences between subjects moderate responses to delays in the feedback during the execution of a task.
% In particular, we view cognition as described by models such as {ACT-R}~\cite{neves1981knowledge}; i.e.\ as a sequence of procedures performed by the subject.

Our overarching theoretical construct is the idea that people perform temporally sequential tasks in the framework of cognition and emotion.
In particular, we view cognition as described by models such as {ACT-R}~\cite{neves1981knowledge}; i.e.\ as a sequence of procedures performed by the subject.
Additionally, we hypothesize that individual diferences between subjects moderate responses to delays in the feedback during the execution of a task.

@@

The question of how people respond to delay in a computer system is grounded in how people perceive time.
Time perception has been described as regulated by an attentional gate that, when opened, starts a cognitive pulse counter~\cite{zakay1995attentional,zakay1996role}.
More recent research indicates, however, that duration perception is highly malleable and the result of multiple timing mechanisms found in overlapping, flexible neural systems~\cite{bruno2016multiple,wiener2011multiple}. 
The estimation of an event's duration varies with context of various types
\begin{enumerate*}[label={(\roman*)}, before=\unskip{: }, itemjoin={{; }}, itemjoin*={{; and }}]
    \item events subsequent to a long or short interval are contracted or extended, respectively~\cite{heron2012duration}
    \item repeated events tend to be perceived as shorter than novel ones~\cite{matthews2011stimulus}
    \item arousal can expand durations~\cite{droit2011emotion}.
\end{enumerate*}

Expectations play a critical role in time perception as well~\cite{zakay1995attentional,zakay1996role}.
It has been shown that people have a general tendency to be hypersensitive to delays in worse-than-expected states, and under-sensitive to meeting or exceeding expectations~\cite{Loewenstein1992anomaliesintertemporalchoice}.
Accordingly, failures to meet expected fast response times tend to be experienced as highly negative, whereas fast latencies are not noticed.
Violations of expectancy have a strong impact on the acceptability of computer systems.
Users of a computer system anticipate the latency for events, for which the standards only become more stringent as systems improve in response time.
In immersive systems like WCA, which aim to provide seamless interaction, delays are particularly noticeable.

It has long been recognized that slow system response times can undermine cognitive processing, slow the pace of users, and lead to stereotyped behavior and errors, as well as cause negative emotional consequences~\cite{dabrowsky:2011:40years}.
However, standards for what constitute tolerable delays have changed dramatically compared to three decades ago, when delays on the order of \SI{10}{\second} were deemed acceptable~\cite{nielsen1994usability, shneiderman2016designing, seow2008designing}.
Today's user context, and WCA in particular, often demand response times orders of magnitude shorter.

For WCA the acceptable range for latencies was explored by Chen et al.~\cite{Chen:AnEmpiricalStudyOfLatency}, by constructing assistants for tasks with a range of time constraints, including step-by-step tasks and more interactive contexts like playing Ping-Pong against a human opponent.
They then proposed a latency tolerance zone according to the task demands.
For an essentially self-paced task like LEGO assembly, they found two key ranges of latency; unnoticeable, \SIrange{0}{0.6}{\second}; and impaired, \SIrange{0.6}{2.7}{\second}. 
Beyond that, users could begin to show the negative outcomes previously catalogued~\cite{dabrowsky:2011:40years}.

@@

A first possible explanation comes from research on cognitive and motor planning. 
Delay may move users from relatively automatic to more attention-demanding processing.
Cognitive and motor tasks are commonly described as a hierarchical system, progressing from high-level goals to the sequence of commands that accomplishes them.
As competency in a task increases, execution of the hierarchy becomes increasingly automated.
Automatization has been described from a computational perspective in {Anderson's ACT-R}~model as the compiling of multiple productions into one~\cite{neves1981knowledge}.
Neural measurements indicate that with automaticity, control moves from frontal brain areas to more posterior ones~\cite{jeon2015degree,puttemans2005changes}, and similar distinctions have been related to temporal processing~\cite{lewis2003distinct,koch2009neural,lee2019limiting}.

Although activities guided by a WCA are not simple motor actions, immediate feedback after each of a series of repeated actions should promote development and automatic execution of a hierarchical plan.
Delays, in contrast, would disrupt such a plan through the loss of automated control~\cite{lee2019limiting}.

A second, alternative explanation of delay effects appeals to emotional systems rather than cognitive processes.
As users of a system become emotionally aroused by delay, they may be subject to generalized arousal, causing decrements in performance~\cite{lee2019limiting}.

Finally, a third potential explanation of delay effects is what has been called ``ego depletion'', the notion that expending effort on self-control eliminates resources needed for further effort~\cite{baumeister74tice,lin2020strong}.

The various processing accounts of delay effects predict different outcomes, which we will consider in the context of the current data.
If delay increases attentional demands on cognitive processes, responses should be slowed and errors expected, particularly on time-critical tasks.
Generalized arousal triggered by emotional stress from delay should emerge in physiological measures, such as increased heart rate or skin conductivity.
Arousal can also reduce movement smoothness or add erratic gestures~\cite{pijpers2003anxiety}.
Ego depletion has been found to produce premature responses culminating in error~\cite{lin2020strong}, or to lead to abandoning a task entirely~\cite{baumeister74tice}.

Over-arching prescriptions for tolerable system response time have not tended to take into account individual differences in users with respect to salient variables like cognitive ability or personality. 
Relevant research can be found in studies of delay discounting, the tendency to devalue rewards for which one must wait.
High discounting rates, indicative of waiting intolerance, have been associated with negative social and academic outcomes.
Hirsh et al.~\cite{hirsh2008delay} found that higher discounting was associated with extraversion among those with low cognitive function, whereas lower discounting was associated with emotional stability (low neuroticism) for people with high cognitive function.
Among computer system users who tend to have relatively high cognitive ability (which presumably describes the present experimental population), this points to neuroticism as a personality factor that might modulate tolerance for waiting. Extraversion could also  be  a moderating factor among the broader target audience of WCA, which are intended for relatively inexperienced users of an application.
These and other measures of individual variation were considered here.

@@

\begin{figure}[h]
  \centering
%   \includegraphics[width=.8\textwidth]{Fig1.eps}
  \caption{Experimental test-bed.
  Participants interact with the cognitive assistant through task-related inputs and outputs --- in practice, these correspond to the video feed captured by the assistant and the instructions provided by it.
  The assistant itself has been instrumented with a data collection layer, which collects and processes experiment-related data such as biometric signals from the participants (these are merely processed here and do not form part of the inputs to the cognitive assistant as such, however), and a delay buffer, which introduces controlled delays in the transit of information from the core processing component.}%
  \label{fig:experimentaltestbed}
\end{figure}

This study was conducted with the approval of the Carnegie Mellon University Institutional Review Board under record number \texttt{STUDY2019\_00000247}. 
Written consent was obtained.

@@

We used a modified version of the LEGO Assistant application introduced by Chen et al.~\cite{Chen:EarlyImplementation}.
This application belongs to a category of WCAs designed to guide users through the execution of a sequential task.
% Such applications constitute ``\emph{conversational computing tasks}'' in the taxonomy proposed by \textcite{dabrowsky:2011:40years}.
The assistant presents a series of instructions to the user in a semi-predetermined order while it monitors the progress in real-time.
Whenever the application detects that the user has correctly performed an instruction, it provides a new one.
In the case of an incorrect action by the user, a procedurally generated corrective instruction is provided instead.
% The results of the user performing these instructions are provided as inputs to the system.
% These inputs may either be correct, in which case the system proceeds to output the next instruction, or incorrect, in which case a procedurally generated corrective step that fixes the mistake is provided to the user.

@@

\begin{figure*}[h]
  \centering
%   \includegraphics[width=.8\textwidth]{Fig2.eps}
  \caption{Example of a Cognitive Assistance task and its component steps.}\label{fig:task:steps}
\end{figure*}

@@

\begin{figure*}[h]
  \centering
  % \input{plots/lego_hierarchy_diagram.tex}
%   \includegraphics[width=\textwidth]{Fig3.eps}
  \caption{Hierarchical cognitive structure of a step in the LEGO task.}
  \label{fig:lego:hierarchical}
\end{figure*}

@@

For the experimental LEGO task, a key modification was made to the structure of the steps.
After each the processing of each input is completed, the result is withheld for a variable period of time until a specific target \emph{delay} is reached, as illustrated in \cref{fig:cogassist:step,fig:cogassist:step:delay}.
The length of this delay is one of two independent variables we manipulated for the experiment.
Seven levels of delay were used --- no added delay (which we will also refer to as \SI{0}{\second} delay), \SIlist{0.6;1.125;1.65;2.175;2.7;3.0}{\second} --- chosen based on the latency bounds found previously by Chen et al.~\cite{Chen:AnEmpiricalStudyOfLatency}.
% A value of  \SI{600}{\milli\second} was identified as the bound where users start noticing delays in the assistant.
% Conversely, \SI{2.7}{\second} was identified as the upper bound on delays after which the application is considered to be in such a degraded state it is basically ``unusable''.
Our selection of delays is centered around the range of delays where latency is noticeable to users, but the application remains in a ``usable'' state.
% while including one delay value in the unnoticeable range and one completely in the ``unusable'' range.

@@

The first of these, the Big Five Inventory of Personality (\emph{BFI}~\cite{john1999:bfi}), consists of 44 questions to be answered on a 5-point \emph{Likert}-type scale, assessing the traits of \emph{agreeableness: detached to compassionate} (8 questions), \emph{conscientiousness: careless to organized} (9 questions), \emph{extroversion: reserved to outgoing} (8 questions), \emph{neuroticism: secure to sensitive} (8 questions), and \emph{openness: cautious to curious} (9 questions).
Of these, extroversion and neuroticism have been related to tolerance for delay~\cite{hirsh2008delay}.
